---
title: "Mitochondria velocity"
author: "Bindoff, A."
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = T)
```


Data from an experiment investigating mitochondria velocity was provided as a flat table by AC by email on 4th September 2017.

Publication standard plots are readily produced once labels and journal style are specified.  

### Statistical Appendix


```{r libs, warning = F}
# install.packages("dplyr")
# install.packages("ggplot2")
# install.packages("lme4")
# install.packages("lmerTest")

library(ggplot2)
library(reshape2)
library(dplyr)
library(lme4)
library(multcomp)
library(lmerTest)

```



#### Exploratory plots

First we need to load the data into an R object. I opened the Excel file then "Save As" a .csv file with an easy to remember name. `read.csv` loads data from a .csv file, using the first row to name the columns. I assign the data to object `df`, then have a look at the first few rows with `head(df)`. The `$` sign tells R to deal with the named part of the `df` object (it is not the only way that we can manipulate objects in R, so it's worth looking at some basic tutorials if this is unfamiliar).


```{r load data}
df <- read.csv("mito_velocity.csv")
head(df)

# where data have come from a spreadsheet (rather than a database) it is good practice to
# be explicit about what type each variable is if it could be interpreted in more than one way
# for example, taxol could be interpreted as an integer, a continuous variable, or a factor with m levels
df$vehicle <- factor(df$vehicle)
df$embryo <- factor(df$embryo)

# column names with spaces get '.' where the spaces were, create a new column and get rid of the old one
df$t.dur <- df$treatment.duration..hours.
df <- df[, -5]

# create a control
df$control <- as.integer(df$taxol == 0 & df$latruculin == 0 & df$nocodazole ==0 & df$jasplakinolide == 0)
df$t.dur[df$control == 1] <- 0
zero2na <- function(x){
  x[x == 0] <- NA
  return(x)
}

df[,c(6,7,8,9,12)] <- apply(df[,c(6,7,8,9,12)], 2, zero2na)

i <- which(df$control == 1)
df[i, c(6,7,8,9)] <- 0


# melt puts the data into 'long' form
# this is important because we want to be able to treat 0 as a control level,
# but if cells were exposed to 0 taxol, but 10 nocodazole - they are not a control

df <- melt(df, id.vars = c("mother", "embryo", "coverslip", "genotype", "t.dur", "velocity", "vehicle", "control"), value.name = "treatment", variable.name = "agent", na.rm = T)

```

The `ggplot2` library makes graphical exploration of your data very easy. The `gg` stands for "grammar of graphics", and it takes a bit of effort to learn this grammar but the package is well documented and there are lots and lots of tutorials aimed at beginners. 

We want 'treatment' to be continuous, numeric data - but if we plot it that way it's not informative so when we call `ggplot` we wrap it in `factor()` so that we can see the treatment levels. We also log transform `velocity` and check that the distribution looks approximately normal.  

```{r plots}
ggplot(df, aes(x = agent, y = log(velocity), fill = factor(treatment))) +
  geom_violin(draw_quantiles = c(0.025, 0.5, 0.975))
```

Cross-tabulate to get an idea of sample sizes

```{r}
with(df, table(treatment, agent))

xtabs(treatment/treatment ~ mother + agent, data = df)

with(df, table(t.dur, treatment))

```

Test main effect of treatment for each agent, conditioning on treatment duration = 2h

```{r mixed effects}
df0 <- filter(df, agent == "taxol", t.dur != 24) %>% mutate(dose = factor(treatment))

m1 <- lmer(log(velocity) ~ dose + (1|mother), df0)
anova(m1)
```
Notice no p-value in the ANOVA table here. This is because there is only one 'mother' for each condition, so the degrees of freedom are zero.  

```{r}
summary(g1s <- glht(m1, linfct=mcp(dose = "Tukey")))
confint(g1s)
```

The Tukey contrasts are still correct, but must be interpreted with caution. The lesson here is to remember that you're working with small numbers of replicates and for each experimental condition you lose degrees of freedom. This means you should strive for balanced experimental designs.


```{r}
df0 <- filter(df, agent == "nocodazole", t.dur != 24) %>% mutate(dose = factor(treatment))

m1 <- lmer(log(velocity) ~ dose + (1|mother), df0)
anova(m1)
summary(g1s <- glht(m1, linfct=mcp(dose = "Tukey")))
confint(g1s)

df0 <- filter(df, agent == "jasplakinolide", t.dur != 24) %>% mutate(dose = factor(treatment))

m1 <- lmer(log(velocity) ~ dose + (1|mother), df0)
anova(m1)
summary(g1s <- glht(m1, linfct=mcp(dose = "Tukey")))
confint(g1s)

df0 <- filter(df, agent == "latruculin", t.dur != 24) %>% mutate(dose = factor(treatment))

m1 <- lmer(log(velocity) ~ dose + (1|mother), df0)
anova(m1)
```

You would report the results along the lines of "a mixed model showed there was no significant effect of dose (F(2, 15.9) = 2.34, p = .128)"


```{r}
summary(g1s <- glht(m1, linfct=mcp(dose = "Tukey")))
confint(g1s)


```

The confidence intervals can probably go into a table (unless you want to figure out how to plot them?), with the p-values reported as "Tukey Contrasts". It's worth searching papers in your field with the terms "mixed model Tukey" to see how others are reporting this sort of result.  
  
  
Test main effect of treatment for each agent, conditioning on treatment duration = 24h

```{r}
df0 <- filter(df, agent == "taxol", t.dur != 2) %>% mutate(dose = factor(treatment))

m1 <- lmer(log(velocity) ~ dose + (1|mother), df0)
anova(m1)
summary(g1s <- glht(m1, linfct=mcp(dose = "Tukey")))
confint(g1s)

df0 <- filter(df, agent == "nocodazole", t.dur != 2) %>% mutate(dose = factor(treatment))

m1 <- lmer(log(velocity) ~ dose + (1|mother), df0)
anova(m1)
summary(g1s <- glht(m1, linfct=mcp(dose = "Tukey")))
confint(g1s)

df0 <- filter(df, agent == "jasplakinolide", t.dur != 2) %>% mutate(dose = factor(treatment))

m1 <- lmer(log(velocity) ~ dose + (1|mother), df0)
anova(m1)
summary(g1s <- glht(m1, linfct=mcp(dose = "Tukey")))
confint(g1s)

df0 <- filter(df, agent == "latruculin", t.dur != 2) %>% mutate(dose = factor(treatment))

m1 <- lmer(log(velocity) ~ dose + (1|mother), df0)
anova(m1)
summary(g1s <- glht(m1, linfct=mcp(dose = "Tukey")))
confint(g1s)

```